---
title: "R3-Gathering-Cleaning"
author: "Katherine Oung"
date: "3/2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Learn how to import a wide variety of files in R including spreadsheets, files on the web and HTML tables, and transform the results into usable data. This session will also focus on how to clean and structure the data you've gathered in preparation for analysis using tidyverse packages.

This session is good for: People who have some experience using R and the Tidyverse.

## Setup

These packages should be installed on your machine already, but for reference, you can one-time install packages in your console using install.packages('packagename').

### Loading packages

```{r}
library(tidyverse)
library(rvest)
library(janitor)
library(readxl)
library(googlesheets4)
library(lubridate)

# Turn off scientific notation
options(scipen=999)
```

## Ways to load data

If you've taken earlier classes in this sequence, you'll have loaded Social Security Administration data on names of babies born from 2000-2024.

Loading data from a flat file -- like a csv -- stored on your local machine is a very common data loading task.  For that, we'll use the `read_csv()` function that's part of the `readr` package that loads with the `tidyverse`. 

You can load data from an absolute or relative (recommended) path on your computer.

#### Loading data from a local flat (csv) file
```{r}
baby_names_flat <- read_csv("data-raw/R3_babynames_2000_2024_NICAR26.csv")

# Show first 10 records
baby_names_flat %>%
  head(10)
```

#### Loading data from a flat file on the Internet

We can load that same csv from the internet, instead of doing it locally.  We just need to change the information inside of the `read_csv()` function to pass a URL, instead of a location on our computer. 

The CSV is up on Github.com, here: https://github.com/karina-kumar/NICAR26-R-Sessions/blob/798ed76f38a0481b8256694fc33bacfbc2bdfd58/data-raw/babynames_2000_2024_NICAR26.csv

Warning: this may not work if the wifi is buggy!

```{r}
baby_names_flat_web <- read_csv("https://raw.githubusercontent.com/karina-kumar/NICAR26-R-Sessions/main/data-raw/R3_babynames_2000_2024_NICAR26.csv")

# Show first 10 records
baby_names_flat_web %>%
  head(10)
```

#### Loading data from an Excel file

Another data format you'll likely see in the wild: Excel files. They're a little trickier than csvs, because they can contain multiple sheets. 

We can use the `read_xlsx()` function from the `readxl` package to load data from Excel files. 

There's an Excel file in the data folder called R3_baby_names_by_letter.xlsx. It is the same as the csv file, but the baby names are seperated into seperate sheets based on the first letter of the name. 

Let's load the first sheet, the "A" babynames, as a dataframe and store it as an object called babynames_a_excel.

Because the "A" names sheet is the first sheet in the Excel file, the function automatically loads the "A" sheet. 

If you want to write a dataframe as an Excel file with multiple sheets, you can also do this with the openxlsx package!

```{r}
baby_names_a_excel <- read_xlsx("data-raw/R3_baby_names_by_letter.xlsx")
```

But if we need to get any other sheet, like the one containing the "Q" names, we'll have to get it by adding another argument to the read_xlsx function. 

The sheet is called "Q".

How can we load Sheet Q?  Let's take a look at the handy help page for the readxl package.

* You can either type ?read_xlsx in the console and look at what pops up on the right. 
* Or you can visit https://readxl.tidyverse.org/reference/read_excel.html.  

Can you see how to read in a different sheet from the first?  

Let's modify the read_xlsx function below to load the sheet "Q".

```{r}
baby_names_q_excel <- read_xlsx("data-raw/R3_baby_names_by_letter.xlsx")
```

#### Loading data from an online HTML table

On this page is an HTML table showing the top 2024 baby names. https://www.ssa.gov/oact/babynames/. Because the wifi is often buggy, I've saved an HTML file locally of this page, which we'll work with instead. It's in the data folder.

We can use the inspector in our web browser to see this.  It's composed of html tags. 

Using the `rvest` package we're going to first read in the entire html page using the `read_html()` function.

```{r}
# If we were going to pull directly from the web, instead of from a local file, this is what it will the function would look

# top_2022_baby_names <- read_html("https://www.ssa.gov/oact/babynames/") 

# Instead, we're gonna load a local html file to simulate the act of scraping from the web
top_2024_baby_names <- read_html("data/html/ssa_baby_names.html")
```

Open it up in the environment window.  It's the full HTML of the page. 
Next, we'll use `rvest` to extract the html table we want using `html_table()`

```{r}
top_2024_baby_names <- read_html("data-raw/ssa_baby_names.html") %>%
  html_table()

```

Open it up in the environment window. We've isolated that one html table and turned it into a dataframe, albeit one that is nested in a list.  Let's extract it from the list. 

```{r}
# Pluck out the table
top_2024_baby_names <- top_2024_baby_names[[1]]

# display it
top_2024_baby_names

```

Never web scraped before? Now you have! :-)

#### Loading data from online Google Sheets

We can also load data directly from Google Sheets, using the `read_sheet()` function from the `googlesheets4` package.  Google Sheets, like Excel files, can also have multiple workbooks. 

I've loaded up a Google Sheet workbook here with two sheets, one for "A" babynames and one for "B" babynames: https://docs.google.com/spreadsheets/d/1GG_RmYPGCNKbLb4x1dkQv1B9NjDBF0kQRtge9Aea4rs. These samples only have 1,000 records in them.

To read in the sheet, we need to paste in the sheet ID, which is the long sting of numbers and letters at the end of the URL: "1GG_RmYPGCNKbLb4x1dkQv1B9NjDBF0kQRtge9Aea4rs"

Because the "A" sheet comes first, it automatically loads that sheet.

Before we get there, though, we'll need to authenticate with Google.  Look down in the console, select an account or hit 1 to start the process of in-browser authentication. Make sure to allow the Tidyverse API to read, edit, and write to your Google Sheets!

```{r}
gs4_auth()

```

Now we can read in the sheet.

```{r}

babynames_a_sheet <- read_sheet("1YfWqUcGlWqBpA2egCP2DHjnBQIB6ed4qlTEoUm7a9mk")
```

If we want to load any other sheet, we need to tell it which one to load. 

```{r}
babynames_b_sheet <- read_sheet("1YfWqUcGlWqBpA2egCP2DHjnBQIB6ed4qlTEoUm7a9mk", sheet = "babynames_b")
```
#### Your Turn

This page https://www.ssa.gov/oact/babynames/state/top5_2021.html, which I've also stored locally, contains tables of the top 5 birth names for 2021 by state, one for male and one for female.  

Using what you just learned, I'd like you to take a stab at ingesting the html of this page, and saving each of the tables as a separate data frame.   

```{r}

# local path to the data 
# "data-raw/baby_names_2021_state.html"



```


### Your Turn Option 2

538 puts a lot of neat data sets on its' GitHub organization page.  Here's the page to a dataset about the Avengers, from Marvel Comics:  

https://github.com/fivethirtyeight/data/blob/master/avengers/avengers.csv

Read this data into your environment and store it as an object called `avengers`. 

```{r}

```


### Other ways of getting data

There are lots of other ways to bring data into R.  We don't have to delve into all of them today, but you should be aware of them for continued learning. 

* Pull data from an SQL database with the `dbplyr` package. [dbplyr](https://dbplyr.tidyverse.org/)

* Get data from an API endpoint - You can access data available through pretty much any API, or application programming interface, by sending queries and getting back a response, using the excellent [HTTR package](https://httr.r-lib.org/).  An example of an API: [Spotify data](https://developer.spotify.com/documentation/web-api/) 

* Get data from packages the R community has built to load and work with specific kinds of data, including some nice helper functions. If the dataset you want to use is popular, there's a great change someone has built an R package for it! Some great examples:
  *[Tidycensus](https://walkerke.github.io/tidycensus/index.html): for loading and working with all manner of U.S. Census Data.
  *[RTweet](https://rtweet.info/): for working with Twitter data

## Data Cleaning

Data cleaning is a critical step in preparing data for analysis.  Everyone wants to jump right to the fun stuff -- me too! -- but failing to account for flaws in the data can lead to errors.  

To start, we're going to load a dataframe with info on year-by-year totals for the most popular female baby names from 2008 to 2017. 

I've intentionally made this very ugly, so we can practice how to clean up our data!

```{r}
babynames_dirty <- read_csv("data-raw/babynames_dirty.csv")
```
### Identifying issues


Examine the data in the environment window.  

What problems do you see? Don't scroll down until you think you've found one or two of the problems. 










Are you sure you can't find more?












Are you sure? Ok. Here's what I see. 

* EMPTY ROWS - There's a totally empty row of NAs. Let's get rid of it.  It could prevent proper summarizing. 
* EMPTY COLUMN - There's a totally empty column called source. Let's get rid of it, because it's unneeded. 
* INCONSISTENT VALUES - The "sex" column has inconsistent values for Female vs F. This could hurt our ability to group and summarize properly. 
* STYLING INCONSISTENCIES - There are lots of issues in "name_at_birth" column. MADISON is upper. elizabeth is lower. Abigail has "First Name: Abigail" instead of just Abigail. Everything else is title case. This is just...ugly.
* DUPLICATE ROWS - There are two apparently identical rows for Emma and Amelia (or are they?). We should really only have one, and could lead to misleading summarization if we kept both.
* NUMBERS STORED AS CHARACTER STRINGS - The 2015 column has a random underscore appended to the end of the number, unlike every other year column. The fact that it's a mix of characters and numbers means the column is stored as a "character", while all of the other year columns are numbers (integers).  If we want to calculate say, the percent change in a name between 2008 and 2015, we'll get an error. 
* BAD COLUMN NAMES - Less obvious, but still annoying: the fact that the year columns start with a number means they'll be harder to include in our code.

Let's examine each of these, while fixing them. 

Our goal will be one nice long chained function to fix all the problems, so we have a clear record of what we've done.


### Remove empty rows and columns

### Remove duplicates

### Cleaning up names columns

### Standardizing

Thanks for joining.  If you get stuck in the future and need some help, I'm at katherine.oung@gmail.com. 

Updated for 2026 with credit to smussenden's NICAR 2024 R3 lesson.


